Computing likelihoods under the SMC for a general class of ARGs.

### 2024-03-14 paper.tex
updates slicing

First, we describe in more detail what sort of information we have in estimated ARGs.
Then, we define the SMC process in a ``backwards-in-time'' way,
rather than the usual ``along-the-genome'' way.
This makes it relatively straightforward to compute the likelihood,
as we explain in the next section.
Finally, we close with a demonstration that the likelihood computation can be used
to do Bayesian inference on ancestral population sizes.

@@

[TO DO: stress fact that this description does not add anything to McVean and Cardin although the SMC is mostly reasoned about in a left to right sense. We use the notation of McVean and Cardin here.]
In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a 
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineage with \emph{overlapping} genomic intervals may coalesce.
Surprisingly, although this is defined in terms of a backwards-in-time Markov process,
the resulting tree-valued process along the genome is also Markov.

@@

% % I don't think we need the notation here:
% The process described in the previous paragraph details the transitions of the 
% set of labeled lineages $L$ through time. An Ancestral Recombination Graph 
% $\mathcal{G} = (N, E)$ capturing the various inheritance paths along the genome 
% is obtained by registering an edge represented by a tuple $(c, p, X_{cp})$ 
% whenever a lineage $c$, associated with a disjoint set of genomic intervals $X_c$, is 
% replaced by one or two lineage(s), $p$ (and $p^{\prime}$), 
% in case of a coalescence or recombination respectively.
% The last element of the tuple is the extent of the edge. It is itself a  
% disjoint set of intervals such that $X_{cp} \cup X_{cp^{\prime}} = X_c$. 
% The set of edges $E$, containing all such tuples $(c, p, X_{cp})$, 
% and $N$, the set of nodes representing all genomes in the ARG, together
% define this data structure.

@@

Furthermore, these unary nodes enable 
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}). 
In the absence of these recombination nodes, a recombination can be observed 
easily enough as a change in parent going from one genomic region to the next.
Without the recombination nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $c$ happened somewhere between node 5 and node 7.
More generally, a recombination node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the 
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently, 
\argweaver{} \citep{rasmussen_genome-wide_2014} 
\tsinfer{} \citep{kelleher_inferring_2019}, and
\kwarg{} \citep{ignatieva_kwarg_2021}
estimate ARGs including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation,kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).
% ARG simplification is thus linked to the presence of unary nodes in 
% the local trees and simultaneously touches upon the question of what is 
% knowable. Many ARG reconstruction methods return a data structure that 
% is in essence a series of local trees [REFs]. 
% As highlighted, this does not need to be a contradiction. 
% All ARGs can be represented as a series of local trees \citep{wong_general_2023}.
% The important question is therefore is however about the extent by which any 
% ARG reconstruction methods incorporates the available information 
% on how nodes persist across these local trees.
% 
% [TO DO] give a good explanation (using the represented haplotypes 
% in fig. \ref{fig:smc-unary}) on why these methods can infer the extent 
% of coalescing lineages.

\comment{[TO DO]: impact of unary nodes on reasoning about the SMC in a left-to-right sense.
Is there more to say them the simple remark I made in \ref{fig:smc-unary}?}
\comment{PLR: Not that I can think of?}

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
\caption{ARG (top) generated under the SMC and its representation as a 
series of local trees (bottom). To ensure the one-to-one correspondance 
between both representations a node is recorded for each local tree whenever 
a node is encoundered along its ancestry path in the graph representation. 
This results in nodes that (locally) have only a single child, and are therefore 
(locally) unary. Blue (unary) nodes represent recombinant nodes. 
The blue arrows shows the familiar left-to-right logic of the SMC whereby 
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
Note how the presence of the unary node on a local tree is 
indicative of either a past (to the left of the local tree) 
or future (to the right) coalescence event. 
Both ARG representations can be simplified by removing the recombinant nodes. 
In that case the information we have on the time to each of these events is 
limited to the windows (blue shaded array) delimited by by the age of 
the child node $c$ associated with that lineage and the minimum age of 
the parents of the edges connecting to $c$. 
    \comment{The bits of genome that leave 5 by the left-hand path are present in 7's diagram but missing from 8's and 9's.}
 [TO DO: wanted to use the 
added in haplotype information to explain why we can actually infer the 
extent of each lineage, and thus create unary nodes in the first place.]
}
\label{fig:smc-unary}
\end{figure}

@@

% ADD IN RESULTS
Figure \ref{sup:fig:vs-argweaver} shows the expected near-perfect correlation 
between the coalescent prior as returned by \argweaver and the likelihood as computed here. 
We simulated 1Mb of data for 10 diploid individuals 
under the SMC using human-like parameters. 100 samples were taking from the posterior every for 
a total of 1000 MCMC iterations.

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge, we can compute a likelihood 
under the SMC for any valid tree sequence. The \tskit library provides a whole range  
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space, 
as well as reduce the number of tracked samples. 
Any such subset of the tree sequence is itself a valid tree sequence 
that can again be used to compute a likelihood with. Our approach therefore strengthens 
the inherent ability of ARGs to provide us with temporal 
resolution on the inferred genealogical history of a sample. 
We can further readily incorporate changes in $N_e$ and $r$, 
both in time as well as along the genome by respectively adding intervals to $I$ 
or by splitting edges when/wherever the parameter set changes.
Figure \ref{fig:3-arg-slices} demonstrates this idea on a toy example. We simulated an ARG with 
two population size changes and subsequently estimated the posterior distribution 
using a uniform prior.

Although we rely heavily on unary nodes to provide us with the necessary 
information on recombination events, our algorithm only implicitly assumes their presence.
Without unary nodes though, our algorithm will detect a recombination event along 
both coalescing lineages in case of a (sub)tree height changes along the ARG as both 
edges would switch parent nodes. This can be mitigated by limiting the number of 
'inferrable' recombination events per coalescence event to 1 (see fig \ref{sup:fig:rec-correction}).

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
\caption{Simulated ARG under Hudson coalescent (left). $N_e$ varied was constant within each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000, (5000:\infty):10 000]$. Remaining parameters: L=10kb, 20 diploid samples, r=2.5e-8, 148 edges in ARG). A single $N_e$ value was estimated for each slice of the ARG as defined by the true population size change time points. The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior. Number of edges per time slice: [68, 57, 23]. Locally unary nodes were omitted for clarity.}
\label{fig:3-arg-slices}
\end{figure}

@@

% a few details to be discussed
[TO DO: polytomies] The above algorithm treats each edge independently which implies 
we can deal with polytomies.
[TO DO: discuss, what if the identity of internal nodes is not preserved across trees]
[TO DO: recombination]: We do not *explicitly* assume a single recombination breakpoint to have 
occurred at each breakpoint.

@@

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal 
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm 
provided here can function as a building block of [...] 
$N_e$-changes as well as multiple populations and migration can be readily incorporated 
in the current algorithm.\\

% 2/ hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the 
very recent past and coalescent simulations for the remaining time are a tried and tested 
approach \citep{bhaskar_distortion_2014, nelson_accounting_2020} to enable large scale 
simulations while avoiding the documented biases of the coalescent relative to the 
Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. The same approach 
could be used for ARG-inference where one can rely on a heuristic inference method for 
the recent past, while using a model-based approach beyond that cutoff point.\\

@@

As a final remark, the described algorithm relies heavily on the presence of locally unary 
nodes in the tree sequence to correctly identify the full extent of all lineages going back 
through time. Although their presence is crucial to our ability embed any local tree in the 
ARG, the concept and their importance are still relatively new (but see 
\citet{wong_general_2023}). Although we have briefly touched upon the observation that in 
the presence of unary nodes we seem to augment the left-to-right logic of the SMC 
with some additional information. 
We believe that the topic of locally unary nodes in general requires further exploration. 
And more specifically, we see it as future work to fully quantify to what extent the 
absence of these locally unary nodes will affect the shape of the likelihood surface given 
any set of parameters.\\

[ TO DO final conclusion?]

>>>>

First, we define the SMC process in a ``backwards-in-time'' way,
rather than the usual ``along-the-genome'' way. 
Note that this is a
reiteration of the original work of \citet{mcvean_approximating_2005}.
Then, we describe
in more detail what sort of information we have in estimated ARGs.
In the subsequent section we connect the two and
show how the ``backwards-in-time'' description
makes it relatively straightforward to compute the likelihood
for a more general class of estimated ARGs.
Finally, we close with a demonstration that the likelihood computation can be used
to do Bayesian inference on ancestral population sizes.

@@

In its backwards-in-time formulation, the SMC \citep{mcvean_approximating_2005} only requires a
simple modification to the coalescent with recombination.
Instead of allowing any pair of lineages to coalesce,
the SMC is the process in which only lineage with \emph{overlapping} genomic intervals may coalesce.
Surprisingly, although this is defined in terms of a backwards-in-time Markov process,
the resulting tree-valued process along the genome is also Markov.

@@

@@

Furthermore, these unary nodes enable
us to uniquely identify all lineages that were hit by a recombination event,
even if we simplify the ARG by removing all recombinant nodes
($a$, $b$, and $c$ in Fig.\ref{fig:smc-unary}).
In the absence of these recombination nodes, a recombination can be observed
easily enough as a change in parent going from one genomic region to the next.
Without the recombination nodes we no longer know the precise time of each recombination event,
but we do still have constraints on the time:
for instance, we know that $c$ happened somewhere between node 5 and node 7.
More generally, a recombination node must happen somewhere between the child node
and the most recent of the parent nodes.
This is again easier to visualise using the
graph representation (blue shaded area, fig.\ref{fig:smc-unary}).
Although these unary nodes are useful and to some degree clearly inferrable,
they are not always represented by simulators or ARG reconstruction methods.
However, to our understanding this is because
the field is used to thinking in terms of the trees
rather than in terms of relationships between haplotypes,
not because of any question of knowability.
Currently,
\argweaver{} \citep{rasmussen_genome-wide_2014}
\tsinfer{} \citep{kelleher_inferring_2019}, and
\kwarg{} \citep{ignatieva_kwarg_2021}
estimate ARGs including unary portions of coalescence nodes.
However, it is unclear to what degree these are well-inferred,
as most evaluation of inferred ARGs thus far has focused on marginal trees
(e.g., \citet{brandt2022evaluation,kelleher_inferring_2019},
but see \citet{deng2021distribution})
and thus completely ignore such unary nodes
(and their implications for inherited haplotypes).

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/smc_custom_2rows_area.png}
\caption{ARG (top) generated under the SMC and its representation as a
series of local trees (bottom). To ensure the one-to-one correspondance
between both representations a node is recorded for each local tree whenever
a node is encoundered along its ancestry path in the graph representation.
This results in nodes that (locally) have only a single child, and are therefore
(locally) unary. Blue (unary) nodes represent recombinant nodes.
The blue arrows shows the familiar left-to-right logic of the SMC whereby
the floating lineage coalesces randomly
with a the remaining portion of the local tree following recombination.
Note how the presence of the unary node on a local tree is
indicative of either a past (to the left of the local tree)
or future (to the right) coalescence event.
Both ARG representations can be simplified by removing the recombinant nodes.
In that case the information we have on the time to each of these events is
limited to the windows (blue shaded array) delimited by by the age of
the child node $c$ associated with that lineage and the minimum age of
the parents of the edges connecting to $c$.
}
\label{fig:smc-unary}
\end{figure}

@@

% ADD IN RESULTS: 
% Figure \ref{sup:fig:vs-argweaver} shows the expected near-perfect correlation
% between the coalescent prior as returned by \argweaver and the likelihood as computed here.
% We simulated 1Mb of data for 10 diploid individuals
% under the SMC using human-like parameters. 100 samples were taking from the posterior every for
% a total of 1000 MCMC iterations.

@@

% slicing of the ARG
By explicitly associating a single likelihood with each edge,

\comment{This makes it sound as if tskit is crucial for this observation.}
we can compute a likelihood
under the SMC for any valid tree sequence. The \tskit library provides a whole range
of subsetting operations on tree sequences allowing us to slice the ARG in time and/or space,
as well as reduce the number of tracked samples.
Any such subset of the tree sequence is itself a valid tree sequence
that can again be used to compute a likelihood with.

Our approach therefore strengthens
the inherent ability of ARGs to provide us with temporal
resolution on the inferred genealogical history of a sample.

Do this independently for any such slice
Although we can readily incorporate changes in $N_e$ and $r$ 
into the likelihood computation, both in time as well as along the genome,

slicing of the ARG can be used to optimize the parameter of interest
for each slice independently.

Figure \ref{fig:3-arg-slices} demonstrates this idea on a toy example.
We simulated an ARG under human-like parameters and with
two population size changes and subsequently estimated the posterior 
distribution using a uniform prior for each slice indepently.

The observation that we can associate a likelihood with any valid tree sequences
urges the need for some caveats. By treating each edge independently we can
naturally deal with the various degrees of completeness to which inheritance
is documented by the different ARG inference methods: e.g.\ polytomies, absence
of locally unary coalescence nodes, unpreserved identify of internal nodes 
across marginal trees, \dots However, these imprecisions do violate the implicit
assumptions of the likelihood computation.
\comment{Do we want these remarks in here, or as part of the Discussion?}
For example, the described algorithm relies heavily on the
presence of locally unary nodes in the tree sequence to correctly identify the 
full extent of all lineages going back through time and therefore recombination
events. Without unary nodes though, our algorithm will detect a recombination
event along both coalescing lineages in case of a (sub)tree height changes along
the ARG as both edges would switch parent nodes. This can be mitigated by limiting
the number of 'inferrable' recombination events per coalescence event to 1 
(see fig \ref{sup:fig:rec-correction}).

\comment{this was previously part of the discussion, have grouped it here.}
Although their presence is crucial to our ability to embed any local tree in the
ARG, the concept and their importance are still relatively new (but see
\citet{wong_general_2023}).
We have briefly touched upon the observation that in
the presence of unary nodes we seem to augment the left-to-right logic of the SMC
with some additional information. Yet, we believe that the topic of locally 
unary nodes in general requires further exploration.
And more specifically, we see it as future work to fully quantify to what 
extent the absence of these locally unary nodes will affect the shape of the
likelihood surface given any set of parameters.

@@

\begin{figure}[!ht]
\centering
\includegraphics[width=\textwidth]{figures/3-slices-whitespace.png}
\caption{Simulated ARG under Hudson coalescent (left). $N_e$ is constant for each of the 3 time slices: $[(0:1000): 10 000, (1000:5000): 5000, (5000:\infty):10 000]$. Remaining parameters: L=10kb, 20 diploid samples, r=2.5e-8, 148 edges in ARG). A single $N_e$ value was estimated for each slice of the ARG as defined by the true population size change time points. The inferred posterior distribution (right) using the SMC-likelihood was obtained using a uniform prior. Number of edges per time slice (from most recent slice to oldest): [68, 57, 23]. Locally unary nodes were omitted for clarity.}
\label{fig:3-arg-slices}
\end{figure}

@@

@@

% 1/ statistical inference
Currently most/all ARG-based inference methods base their analysis on marginal
trees \citep{hejase_2022} [TO DO: add more refs]. The theory and associated algorithm
provided here can function as a building block of [...]
$N_e$-changes as well as multiple populations and migration can be readily incorporated
in the current algorithm.

% 2/ hybrid ARG-inference: scalability + bringing ARG down from pedigree dominated phase
Hybrid backwards-in-time simulations that combine Wright-Fisher dynamics in the
very recent past and coalescent simulations for the remaining time are a tried and tested
approach \citep{bhaskar_distortion_2014, nelson_accounting_2020} to enable large scale
simulations while avoiding the documented biases of the coalescent relative to the
Wright-Fisher model \citep{bhaskar_distortion_2014, wakeley_gene_2012}. The same approach
could be used for ARG-inference where one can rely on a heuristic inference method for
the recent past, while using a model-based approach beyond that cutoff point.

@@
